Worflow

### 2017-06-08 R/workflow.Rmd
slingshot wi and wo zinbwave

---
title: "Worflow"
author: "Workflow team"
date: "06/06/2017"
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    code_folding: hide
    toc_float: yes
---

@@

FP: yes, I agree.

@@

# TL;DR

This document is a temptative of workflow with the following steps

@@

# Steps of the workflow
## Create a SummarizedExperiment object

@@

In published work (FP: add ref), cells have been clustered into `r length(unique(clus.labels))` different clusters

@@

Let's create a SummarizedExperiment object to store the raw counts and information about the data, that is batches, published labels, and quality control measures.

@@

## Dimensionality reduction adjusting for gene and cell level covariates

@@

### W

Let's check that performing MDS on W we have something coherent with published clusters.

@@

## Cluster of cells

@@

```{r rsec_50}
W <- colData(se)[, grepl('^W', colnames(colData(se)))]
W <- as.matrix(W)
fn <- '../data/RSEC_W.rda'
if (runClus & !file.exists(fn)){
	metadata$clusterLabels[is.na(metadata$clusterLabels)]<- -2 #symbol for samples missing from original clustering
  seObj<-SummarizedExperiment(t(W),colData=metadata)
 print(system.time(  ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1), betas = 0.8,
                clusterFunction = "hierarchical01", minSizes=1,
                ncores = NCORES, isCount=FALSE,
                subsampleArgs = list(resamp.num=100,
                                     clusterFunction="kmeans",
                                     clusterArgs=list(nstart=10)),
                seqArgs = list(k.min=3, top.can=5), verbose=TRUE,
                combineProportion = 0.7,
                mergeMethod = "none")
				))
  save(ceObj, file = fn)
}else{
  load(fn)
}
```

```{r examineCombineMany}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(200)))
```

@@

@@

@@

```{r compareClusters}
plot(fit$points, col = col_clus[as.character(clus.labels)],
     main = 'MDS W, color = published clusters', pch = 20,
     xlab = 'Component1', ylab = 'Component2')
legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = 'Sample')
```

```{r}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, 'color']
names(pal) <- palDF[, 'name']
plot(fit$points, col = pal[primaryClusterNamed(ceObj)],
     main = 'MDS W, color = our new clusters', pch = 20,
     xlab = 'Component1', ylab = 'Component2')
legend(x = 'bottomright', legend = names(pal), cex = .5,
       fill = pal, title = 'Sample')
```

@@

```{r}
# sampleData <- data.frame(ours = primaryCluster(ceObj))
# plotHeatmap(assays(se)$normalizedValues,
#             main = 'Normalized values, 1000 most variable genes',
#             clusterSamplesData = ceObj@dendro_samples,
#             sampleData = as.matrix(sampleData),clusterLegend=ceObj@clusterLegend[1])
# easier fix:
plotHeatmap(ceObj, visualizeData=assays(se)$normalizedValues,
whichClusters="primary",
            main = 'Normalized values, 1000 most variable genes')

@@

## Get lineages

```{r}
table(data.frame(published = clus.labels,
                 ours = primaryClusterNamed(ceObj)))
```

Cluster name | Description | Correspondence
-------------|-------------|-----------------
m1 | resting HBC | original 1, 5
m2 | activated HBC | original 5
m3 | GBC / immature neurons / MV 1 | original 2, 3, 11
m4 | Sus | original 4, 7
m5 | new | new
m6 | new | new
m7 | Neuron | original 9, 12
m8 | Immature Neuron | original 10, 14
m9 | Microvillus | original 15
m10 | Immature Neuron | original 14

From what I understand, start original cluster are 1 and 5 (HBC) and end original clusters are 15 (Microvillus), 9 and 12 (neuron), and 4, 7 (Sus).

The goal of this section is to see if we need to refit zinbwave when we want to run slingshot. We run slingshot in the supervised and unsupervised mode and we try k=3, k=4 in W.

@@

```{r}
Kvec <- c(3, 4)
```

### Use previous W, no re-fitting 
#### Unsupervised

```{r slingshot_unsup}
our_cl <- as.factor(primaryClusterNamed(ceObj))
cl = our_cl[our_cl != "-1"]
for (k in Kvec){
  X <- W[our_cl != "-1", 1:k]

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1")
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
}
```

@@

```{r slingshot_sup}
# m11 - original 15 - Microvillus
# m7, m8 - original 9, 12 - Neuron
# m4 - original 4, 7 - Sus

for (k in Kvec){
  X <- W[our_cl != "-1", 1:k]

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1",
                           end.clus = c("m11", "m7", "m8", "m4"))
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
}
```

@@

@@

```{r slingshot_unsup_refit}
for(k in Kvec) {
  X <- getW(zinbList[[k - 1]])[, 1:k]

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1")
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
}
```

@@

```{r slingshot_sup_refit}
for(k in Kvec){
  X <- getW(zinbList[[k - 1]])[, 1:k]

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1",
                           end.clus = c("m11", "m4"))
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
}
```

## DE analysis

Here is the kind of plots we want to be able to have

```{r}
de <- read.csv('../data/oe_markers.txt', 
              stringsAsFactors = F, header = F)
de <- de$V1
nv <- assays(se)$normalizedValues
plotHeatmap(nv[rownames(nv) %in% de, ],
            main = 'Normalized values, 1000 most variable genes',
            clusterSamplesData = ceObj@dendro_samples,
            sampleData = as.matrix(sampleData$ours))
```

>>>>

---
title: "Worflow"
author: "Workflow team"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    code_folding: hide
    toc_float: yes
---

@@

FP: yes, I agree. Davide, we wait for zinbwave to be on bioconductor before to push changes, right?

@@

# Steps of the workflow

We propose a worklow to analyze single cell RNA-Seq with the following steps

@@

```{r frog-picture, out.width='90%', fig.align='center', fig.cap = 'Worflow to analyze single cell RNASeq data'}
knitr::include_graphics('../vignettes/workflow.png')
```

## 1. Create a SummarizedExperiment object

@@

In original work (FP: add ref), cells have been clustered into `r length(unique(clus.labels))` different clusters

@@

Let's create a SummarizedExperiment object to store the raw counts and information about the data, that is batches, original labels, and quality control measures.

@@

## 2. Dimensionality reduction adjusting for gene and cell-level covariates

@@

### Dimensionality reduction

Let's check that performing MDS on W we have something coherent with original clusters.

@@

## 2. Clustering of the cells

@@

```{r rsec_50}
W <- colData(se)[, grepl('^W', colnames(colData(se)))]
W <- as.matrix(W)
fn <- '../data/RSEC_W.rda'
if (runClus & !file.exists(fn)){
  #symbol for samples missing from original clustering
	metadata$clusterLabels[is.na(metadata$clusterLabels)]<- -2 
  seObj<-SummarizedExperiment(t(W),colData=metadata)
  print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1),
                                  betas = 0.8,
                clusterFunction = "hierarchical01", minSizes=1,
                ncores = NCORES, isCount=FALSE,
                subsampleArgs = list(resamp.num=100,
                                     clusterFunction="kmeans",
                                     clusterArgs=list(nstart=10)),
                seqArgs = list(k.min=3, top.can=5), verbose=TRUE,
                combineProportion = 0.7,
                mergeMethod = "none")))
  save(ceObj, file = fn)
}else{
  load(fn)
}
```

```{r examineCombineMany}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
```

@@

FP: Ha ok, good to know. I'll keep the eval=FALSE for the moment.

@@

FP: Not sure we should show the following heatmap using the W, we should rather use the normalized values (following heatmap).

@@

@@

FP: it seems great to me, what do you think?

```{r}
# sampleData <- data.frame(ours = primaryCluster(ceObj))
# plotHeatmap(assays(se)$normalizedValues,
#             main = 'Normalized values, 1000 most variable genes',
#             clusterSamplesData = ceObj@dendro_samples,
#             sampleData = as.matrix(sampleData),clusterLegend=ceObj@clusterLegend[1])
# easier fix:
plotHeatmap(ceObj, visualizeData = assays(se)$normalizedValues,
            whichClusters = "primary",
            clusterSamplesData = "dendrogramValue",
            main = 'Normalized values, 1000 most variable genes',
            breaks = 0.99)

@@

```{r compareClusters}
plot(fit$points, col = col_clus[as.character(clus.labels)],
     main = 'MDS W, color = original clusters', pch = 20,
     xlab = 'Component1', ylab = 'Component2')
legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = 'Sample')
```

```{r}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, 'color']
names(pal) <- palDF[, 'name']
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)],
     main = 'MDS W, color = our new clusters', pch = 20,
     xlab = 'Component1', ylab = 'Component2')
legend(x = 'bottomright', legend = names(pal), cex = .5,
       fill = pal, title = 'Sample')
```

## 4. Pseudotime ordering

The goal of this section is to see if we need to refit zinbwave when we want to run slingshot. We first run slingshot on the W used by clusterExperiment. In the second part of this section, we fit zinbwave on the matrix of counts where the unassigned cells have been removed. For each part (without or with refitting zinbwave), we run slingshot in the supervised and unsupervised mode and try k=3, k=4, k=5 dimensions in W.

From what I understand, start original clusters are 1 and 5 (HBC) and end original clusters are 15 (Microvillus), 9 and 12 (neuron), and 4, 7 (Sus). Additionally, we want the GBC cluster to be a junction before the differentiation between Microvillus and Neuron. The correspondance with the original clusters is as follow

```{r}
table(data.frame(original = clus.labels, ours = primaryClusterNamed(ceObj)))
```

Cluster name | Description | Correspondence | Color
-------------|-------------|-----------------|---------
c1 | HBC | original 1, 5 | darkblue
c2 | new and small | new and small | green
c3 | new and small | new and small | red
c4 | GBC / immature neurons / MV 1 | original 2, 3, 11, 14 | orange
c5 | Sus | original 4, 7 | darkpurple
c6 | Neuron | original 9, 12 | brown
c7 | Immature Neuron | original 10, 14 | cian 
c8 | Immature Neuron | original 14 | lightpurple
c9 | Microvillus | original 15 | grayblue

@@

FP: With all the improvements made, I think we simply have the correct colors now.

```{r}
Kvec <- c(3, 4, 5)
```

### Use previous W

The input of slingshot is the W used for clusterExperiment where the number of dimensions is reduced to k where k in (3, 4, 5) here.

#### Unsupervised

K = 3 does not seem very good to me: Sus is not an end cluster, GBC is an end cluster.

K = 4 is better, slingshot finds the end clusters but there is a spurious end cluster.

K = 5 does not seem great to me: GBC is an end cluster and Sus and Microvillus are in the same lineage.

```{r slingshot_unsup}
our_cl <- primaryClusterNamed(ceObj)
cl = our_cl[our_cl != "-1"]
pal = pal[names(pal) != '-1']
```

```{r}
for (k in Kvec){
  X <- W[our_cl != "-1", 1:k]

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
  print(lineages$lineage5)
}
```

@@

K = 3 finds GBC as an end cluster (that I did not specify in the end.clus!).

K = 4 Yeah! it seems that it is what we want even if we still have a spurius end cluster and GBC not really at the junction.

K = 5 Yeah! even if GBC not really at the junction.

```{r slingshot_sup}
for (k in Kvec){
  X <- W[our_cl != "-1", 1:k]

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
                           end.clus = c("c5", "c6", "c9"))
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
  print(lineages$lineage5)
}
```

@@

K = 3 is better than when we did not refit zinbwave but still not perfect: Sus in all the clusters. GBC not really at the junction.

K = 4 good even if GBC not really at the junction.

K = 5 not great GBC is an end cluster.

@@

```{r slingshot_unsup_refit}
for(k in Kvec) {
  X <- getW(zinbList[[k - 2]])[, 1:k]

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
  print(lineages$lineage5)
}
```

@@

K = 3 Yeah! close to perfection.

K = 4 good even if GBC not really at the junction.

K = 5 no, GBC is an end cluster.

```{r slingshot_sup_refit}
for(k in Kvec){
  X <- getW(zinbList[[k - 2]])[, 1:k]

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1",
                           end.clus = c("c5", "c6", "c9"))
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
  print(lineages$lineage5)
}
```

CONCLUSION: K = 5 is never great as GBC is generally an end cluster. K = 4 is ok for all the methods and a bit better when zinbwave is refitted. K = 3 when refitting and supervized is good.

It seems to me that using slingshot on W without re-fitting zinbwave with k = 4 gives good results where supervized mode is slightly better than unsupervized. It is just a one shot example and we should obviously not make a general conclusion, but I think that for the purpose of the workflow it is fine to use slingshot without refitting zinbwave. We should write a note to the user that it is better to refit zinbwave to have more power.

## 5. DE analysis

Here is the kind of plots we want to present

```{r}
de <- read.csv('../data/oe_markers.txt', stringsAsFactors = F, header = F)
de <- de$V1
plotHeatmap(ceObj, 
            visualizeData = assays(se)$normalizedValues[rownames(se) %in% de, ],
            clusterSamplesData = "dendrogramValue",
            whichClusters = "primary",
            main = 'Normalized values, 1000 most variable genes',
            breaks = 0.99)
```

FP: Kelly, is it you who did the DE analysis for Russell paper? If yes, what tool did you use? On what data? The full quantile normalized counts? Do you have code available?
